---
title: "Purrr"
author: "Miguel Arias"
date: "9/12/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=6, fig.height=4)
```

## Purrr

```{r start, include = FALSE}
library(purrr)
ghr <- repurrrsive::gh_repos
```

`ghr` is a massive multi-level list that is made up of 6 lists, each representing a GitHub **user**. Each of these lists is made up of around 30 smaller lists and each list has +60 fields with **info** about the repo. 

Let's see how many repositories the first user in `ghr` have:
```{r user1}
length(ghr[[1]])

# If we wanted to see how many information fields are  
# there for the 3rd repo owned by this user:
length(ghr[[1]][[3]])
```

In order to find how many repos each user owns:
```{r loop1}
lengths <- c()
for(i in seq_along(ghr)) {
  lengths <- c(lengths, length(ghr[[i]]))
}
lengths
```

Simpler way! Use `map()`
```{r map1}
# List format
map(ghr, length)

# Double format
map_dbl(ghr, length)
```

### Anonymous functions

Formulas are preceded by a tilde "~", can't control what their arguments are named. Functions are regular R functions.

```{r map2}
map_dbl(ghr, ~length(.x))
```

By using the tilde notation, we can *explicitly* access that element and place it wherever we want inside the function call, but it's name is going to be `.x` no matter what.

```{r map3}
map(1:3, ~runif(2, max = .x))
```

`map()` conveniently allows you to send any fixed arguments you want at the end of the call (note how this time `1:3` is used automatically as the first argument):

```{r map4}
map(1:3, runif, min = 3, max = 6)
```

### Mapping the map

Can next `map()` calls. Useful when accessing deeper levels of a list. Let's see how many information fields each repo of each user has:

```{r map5}
map(ghr, ~map(.x, length))
map(ghr, ~map_dbl(.x, length))
```

The first one outputs a very long list of lists, while with `map_dbl()` on the innermost call allows us to output a single vector for each user.

Let's see if we can find the names of the users in the `login` field of the `$owner` list of each repo (note the use of `map_chr()`; this is the character equivalent of `map_dbl()`):

```{r map6}
# First
map(ghr, function(user) {
  map_chr(user, ~.x$owner$login)
})

# Second
map(ghr, function(user) {
  user %>% map_chr(~.x$owner$login)
})

# Third
map(ghr, ~map_chr(.x, ~.x$owner$login))
```

Only once for every user:
```{r map7}
map_chr(ghr, ~map_chr(.x, ~.x$owner$login)[1])
```

### Set names

Use mapping to ge the logins of the users, then `set_names()` to name the user lists according to their logins and finally `pluck()` the list of repos belonging to "jennybc". The dot in `set_names()` stands for the results incoming from the line above:

```{r map8}
ghr %>%
  map_chr(~map_chr(.x, ~.x$owner$login)[1]) %>%
  set_names(ghr, .)
  pluck("jennybc")
```

To name the second level (repos). Need to go a level deeper and set the names there too:

```{r map9}
ghr %>%
  map(function(user) {
    user %>%
      set_names(map(., ~.x$name))
  }) %>%
  pluck("jennybc", "eigencoder")

ghr %>%
  map(~set_names(.x, map(.x, ~.x$name))) %>%
  pluck("jennybc", "eigencoder")
```

### Other map functions

* `map2()`: iterates over two lists instead of only note ( note that for formulas we'd use `.x` for the elements of the first list and `.y` for the elements of the second).
* `map2_xxx()`: analogous to `map_xxx()`
* `pmap()`: can map over however many elements you want
* `lmap()`: for mapping with functions that take and return lists
* `imap()`: for mapping over a list and its names, just like we did above
* `map_at()/map_if()`: functions that allow you to filter which elements get mapped.