---
title: "Purrr - Wickham"
author: "Miguel Arias"
date: "9/13/2017"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Cwickmham Purrr-Tutorial

### Chapter 1 - map
```{r start, include = FALSE}
library(purrr)
library(repurrrsive)
library(tidyverse)
```

Includes objects: sw_films, sw_people, sw_vehicles, sw_starships, sw_planets & sw_species

How many elements are in sw_people
```{r people1}
length(sw_people)
```

Who is the first person in sw_people?
```{r people2}
sw_people[[1]] # Luke Skywalker
```

To obtain the first person and it's name:
```{r people3}
sw_people[[1]][1]

# A list inside a list
sw_people[1]

# A list, dropped one level of hierarchy
sw_people[[1]]
```

### map()

`map( .x, .f, ...)` for each element of `.x` do `.f`.

* `.x`: a vector, a list, a data frame (for each column)
* `.f`: 

How many starships has each character been on?

1. Do it for one element
```{r one_element}
luke <- sw_people[[1]]
length(luke$starships)
```

2. Turn it into a recipe
```{r recipe1}
# ~ length( .x$starships)
```

3. Use `map()` to do it for all elements
```{r all_elements}
starships <- map(sw_people, ~length(.x$starships))
print(starships[0:5])
```

For later
```{r planet_lookup}
planet_lookup <- map_chr(sw_planets, "name") %>%
  set_names(map_chr(sw_planets, "url"))
planet_lookup
# save(planet_lookup, file = "Purrr/data/planet_lookup.rda", compress = FALSE)
```

Find the name of each characters home world
```{r homeworld1}
luke$homeworld
planet_lookup[luke$homeworld]
map(sw_people, ~ planet_lookup[.x$homeworld])
```

### ROADmap()

`map()` always returns a list:

* Other types of output
+ `map_lgl()`: logical vector
+ `map_int()`: integer vector
+ `map_dbl()`: double vector
+ `map_chr()`: character vector
+ `walk()`: when you want nothing at all, use a funciton for its side effects

```{r names_useful}
# names can be useful
sw_people <- sw_people %>%
  set_names(map_chr(sw_people, "name"))
```

**REPLACE** `map()` with the appropriately typed function
```{r replace}
# How many starships has each character been in?
map_int(sw_people, ~ length(.x[["starships"]]))

# What color is each character's hair?
map_chr(sw_people, ~.x[["hair_color"]])

# Is the character male?
map_lgl(sw_people, ~.x[["gender"]] == "male")

# How heavy is each character?
map_chr(sw_people, ~ .x[["mass"]]) %>% # Want is as a number
  readr::parse_number(na = "unknown")
```

* Other ways of specifying `.f`
+ `.f` can be a formula: `map(.x, .f = ~ DO SOMETHING WITH .x)`
+ `.f` can be a string or integer:
`map(.x, .f = "some_name")`
`map(.x, ~ .x[["some_name"]])`
+ `.f` can be a function:
`map(.x, .f = some_function, ...)`
`map(.x, ~ some_function(.x, ...))`

### Star Wars Challenges

1. Which film (see `sw_films`) has the most characters?
```{r sw_challenge1}
# Set names of the films
# sw_films1 <- sw_films %>%
#   set_names(map_chr(sw_films, "name"))

# For the first film
length(sw_films[[1]]$characters) 

# Number of characters for all films
map_int(sw_films, ~length(.x$characters)) %>%
  set_names(map_chr(sw_films, "title")) %>% # Add "titles"
  sort(decreasing = TRUE)

# Cwickham version - Shorter (more understandable)
map(sw_films, "characters") %>%
  map_int(length) %>%
  set_names(map_chr(sw_films, "title")) %>%
  sort(decreasing = TRUE)
```

2. Which `sw_species` has the most possible eye colors?
```{r sw_challenge2}
# Do it for one
sapply(gregexpr("\\W+", sw_species[[1]]$eye_colors), length) + 1

# Do it for all
map_dbl(sw_species, ~ sapply(gregexpr("\\W+", .x$eye_colors), length) + 1) %>%
  set_names(map_chr(sw_species, "name")) %>%
  sort(decreasing = TRUE)

# CWickham Version - No need for sapply
map_chr(sw_species, "eye_colors") %>%
  strsplit(", ") %>%
  map_int(length) %>%
  set_names(map_chr(sw_species, "name")) %>%
  sort(decreasing = TRUE)
```

3. Which `sw_planets` do we know the least about (i.e. have the most `"unknown"` entries)?
```{r sw_challenge3}
# For one (Checks if there are unknown[NA] values)
map_lgl(sw_planets[[61]], ~ "unknown" %in% .x) %>%
  sum()

# For all
map_int(sw_planets,
        ~ map_lgl(.x, ~ "unknown" %in% .x) %>% sum()) %>%
  set_names(map_chr(sw_planets, "name")) %>%
  sort(decreasing = TRUE)
```

### Iteration functions

`map2(.x, .y, .f, ...)`

`map2_xxx()` works the same way as `map_xxx()`.

* For each function, which two arguments might be useful to iterate over?
+ `download.file()`: for each url download to `destflie`. `walk2()`, `map2_int()`.
+ `rnorm()`: for each n generate a Normal sample with `mean` (or `sd`).
+ `lm()` for each data fit a model (`formula`)
+ `predict.lm()` for each model(object), generate predictions at data (`newdata`)
+ `readr::write_csv()` for each data frame (x) save to path. Similar for `ggplot::ggsave()` for each `plot` save to `filename`. `walk2()`

1. Use `map2()`. For each country in `gap_split_small` create a ggplot of life expectancy through time with a title.

```{r life_expectancy}
# Data
gap_split_small <- gap_split[1:10]
countries <- names(gap_split_small)

# Do it for one
ggplot(data = gap_split_small[[1]], 
       mapping = aes(x = year, y = lifeExp)) +
  geom_point(aes(size = pop)) +
  geom_smooth(se = FALSE) +
  labs(
    title = countries[[1]],
    x = "Time",
    y = "Life Expectancy"
  ) +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))

# Create a function for Life Expectancy over time
countries_lifexp <- function(data, name) {
  ggplot(data, aes(year, lifeExp)) +
       geom_point(aes(size=pop)) +
       geom_smooth(se = FALSE) +
       labs(
         title = name,
         x = "Time",
         y = "Life Expectancy"
       ) +
       theme_bw() + 
       theme(plot.title = element_text(hjust = 0.5))
}

# For all the countires
plots <- map2(gap_split_small, countries, ~countries_lifexp(.x, .y))

plots[[3]]

# In order to print all plots:

# walk(plots, print)
```

### Purrr and list columns

**Data** should be in a data frame as soon as it makes sense!
Data frame: **cases** in rows, **variables** in columns

#### Purr can help turn lists into tibbles

```{r lists_tibble}
library(tidyverse)
library(repurrrsive)
# A useful lookup table -----------------------------------------------
film_number_lookup <- map_chr(sw_films, "url") %>%
  map(~ stringr::str_split_fixed(.x, "/", 7)[, 6])  %>%
  as.numeric() %>%
  set_names(map_chr(sw_films, "url"))

# Use purrr to turn list into table
people_tbl <- tibble(
  name    = sw_people %>% map_chr("name"),
  films   = sw_people %>% map("films"), # results in list column
  height  = sw_people %>% map_chr("height") %>%
    readr::parse_number(na = "unknown"), # needs some parsing
  species = sw_people %>% map_chr("species", .null = NA_character_) # isn't in every element
)

# Turning parts of our list to a tibble ---------------------------------
people_tbl$films

# Use map with mutate to manipulate list columns
people_tbl <- people_tbl %>%
  mutate(
    film_numbers = map(films,
      ~ film_number_lookup[.x]),
    n_films = map_int(films, length)
  )

people_tbl %>% dplyr::select(name, film_numbers, n_films)
```

Create a new character column that collapses the film numbers into a single string,

e.g. for Luke: "6,3,2,1,7"

```{r collapse_num}
# Apply paste0 function to create character vector
people_tbl <- people_tbl %>%
  mutate(films_squashed = map_chr(film_numbers, paste0,
                                  collapse = ", "))

people_tbl %>% dplyr::select(name, n_films, films_squashed)
```

### Other features of purrr

* `pmap( .l, .f, ...)`: to each element of each vector in `.l`, apply `.f`.
* `invoke_map(.f, .x, ...)`: for each function in `.f`, apply it to `.x`

### Lists and functions

`purrr` provides a pile of functions to make working with them easier
**Functions**: `safely()`, `possibly()`, `partial()`.
**Lists**: `transpose()`, `accumulate()`, `reduce()`, `every()`, `order_by()`.

