---
title: "Core Purrr - str() and listviewer()"
author: "Miguel Arias"
date: "9/13/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(repurrrsive)
```

## Str() and Listviewer()

### Inspect and explore

List inspection is very important.

```{r listviewer}
library(listviewer)
```

Before applying a function to every element of a list, better understand it.

Need to know:
* The length of the list
* Are the components homogeneous, i.e., do they have the same overall structure, albeit containing different data?
* Note the length, names, and types of the consituent objects

`str()` helps with basic list inspection. 
* By using `max.level` and `list.len` args can maintain the output of `str()` to manageable volume.

#### listviewer

Often more helpful to explore such lists with an interactive widger, which we include below for examples. 

### Use of `str()`

`got_chars` is a list with informaiton on the 29 point-of-view characters from the first five books in the Song of Ice and Fire series. 
```{r GOT}
str(got_chars, list.len = 3)
str(got_chars[[1]], list.len = 8)
```

`gh_users` is a list with information on 6 GitHub users. `gh_repos` is a nested list, also of length 6, where each component is another list with information on up to 30 of that user's repositories.
```{r github}
str(gh_users, max.level = 1)
```