---
title: "jennybc_purrr basics"
author: "Miguel Arias"
date: "9/13/2017"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Vectors and Lists

### Atomic Vectors

Useful to understand lists as a data structure that generalizes atomic vectors.

The garden variety R object is an **atomic vector** like these:

```{r atomic_vector}
(v_log <- c(TRUE, FALSE, FALSE, TRUE))
(v_int <- 1:4)
(v_doub <- 1:4 * 1.2)
(v_char <- letters[1:4])
```

Each **atom** has the same flavor, by which I roughly mean type or storage mode, and is a scalar, which means "has length one". 

##### Exercises

1. Define the vectors above or similar. Use the family of `is.*()` functions to confirm vector type, e.g. `is.logical()`.

```{r vector_type1}
is.logical(v_log)
is.integer(v_int)
is.double(v_doub)
is.character(v_char)
```

2. What do `is.numeric()`, `is.integer()`, and `is.double()` return for the vectors that hold floating point number versus integers?

```{r vector_type2}
is.numeric(v_doub) # Returns logical value TRUE
is.integer(v_doub) # Returns logical value FALSE
is.double(v_doub)  # Returns logical value TRUE
```

To "index a vector" means to address specific elements or atoms, either for reading or writing. We index a vector using square brackes, like so: `x[something]`. 
* Logical vector: keep elements of `x` for which `something` is `TRUE` or drop those for which it's `FALSE`

```{r logical_vector}
v_char[c(FALSE, FALSE, TRUE, TRUE)]
v_char[v_log]
```

* Integer vector, all positive: the elements specified in `something` are kept
* negative integers, all negative: the elements specified in `something` are dropped

```{r integer_vector}
v_doub[2:3]
v_char[-4]
```

* Character vector: presumes that `x` is a named vector and the elements whose names are specified in `something` are kept *not shown here, since none of the vectors are named*.

### Coercion

There's a hierechy of types: the more primitive ones cheerfully and silently convert to those higher up in the food chain. Here's the order:
1. logical
2. integer
3. double
4. character

For explicit coercion, use the `as.*()` functions

```{r coercion1}
v_log
as.integer(v_log)
v_int
as.numeric(v_int)
v_doub
as.character(v_doub)
as.character(as.numeric(as.integer(v_log)))
```

**Coercion** can be triggered by other actions, such as assigning a scalar of the wrong type into an existing vector. 

```{r coercion2}
v_doub_copy <- v_doub
str(v_doub_copy)

v_doub_copy[3] <- "uhoh"
srt(v_doub_copy)
```

### Lists

Lists useful when:
1. Individual atoms might have length greater than 1.
2. Individual atoms might not be of the same flavor.

**A list is actually a vector in R, but it's not an atomic vector**. We construct a list explicitly with `list()` but, like atomic vectors, more lists are created some other way in real life.

```{r lists1}
(x <- list(1:3, c("four", "five")))
(y <- list(logical = TRUE, integer = 4L, double = 4 * 1.2, character = "character"))
(z <- list(letters[26:22], transcendental = c(pi, exp(1)), f = function(x) x^2))
```

Lists components can:
1. Be heterogeneous
2. Have different lengths
3. Have names, or not

Lists are more general atomic vectors, but just like atomic vectors they have length and can be indexed.

### List indexing

Three ways:

1. With single square brackets, i.e. just like we indexed atomic vectors. Note this **always returns a list**, even if we request a single component.

```{r list_index1}
x[c(FALSE, TRUE)]
y[2:3]
z["transcendental"]
```

2. With double square brackets. This can only be used to access a single component and it returns the "naked" component. You can request a compoennt with a positive integer or by name.

```{r list_index2}
x[[2]]
y[["double"]]
```

3. With the `$`, which you may already use to extract a single variable from a data frame (which is a special kind of list!). Like `[[`, this can only be used to access a single component, but it is even more limited: Must specify the component by name

```{r list_index3}
z$transcendental
```

### Vectorized operations

Consider how to square the integers 1 through `n`. 

```{r vect_op1}
# Brute force method
n <- 5
res <- rep(NA_integer_, n)
for(i in seq_len(n)) {
  res[i] <- i ^2
}
res

# R way
n <- 5
seq_len(n) ^ 2
```

However, some functions do not work with lists.

In order to apply the functions to a list:
* Use `purrr:map()`. The first arg is the list to operate on. The second is the function to apply.

```{r vect_op2}
library(purrr)
l_doub <- as.list(v_doub)
exp(v_doub)
map(l_doub, exp)
```
