---
title: "Chapter 12"
author: "Miguel Arias"
date: "9/5/2017"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Tidy data

```{r start, echo = FALSE}
library(tidyverse)
```

Examples of data organised in four different ways.

```{r tidydataorg}
table1
table2
table3
table4a
table4b
```

All representations of the same underlying data, but they are not equally easy to use. There are three interrelated rules which make a dataset tidy:

1. Each variable must have its own column
2. Each observation must have its own row
3. Each value must have its own cell

These rules lead to an even simpler set of practical instructions:

1. Put each dataset in a tibble.
2. Put each variable in a column.

In this example, only `table1` is tidy.

Tidy data facilitates transforming it.

```{r tidydata1}
# Compute rate per 10,000
table1 %>%
  mutate(rate = cases / population * 10000)

# Compute cases per year
table1 %>%
  count(year, wt = cases)

# Visualise changes over time
library(ggplot2)
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```

## 12.3 Spreading and gathering

**Gather()**

```{r gather1}
table4a
```

Need to `gather()` the columns into a new pair of variables.

```{r gather2}
table4a %>%
  gather(`1999`, `2000`, key = "year", value = "cases")
```

```{r gather3}
table4b %>%
  gather(`1999`, `2000`, key = "year", value = "population")
```

To combine the tidied versions of `table4a` and `table4b` into a single tibble, we need to use `dplyr::left_join()`.

```{r gather4}
tidy4a <- table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")
tidy4b <- table4b %>% 
  gather(`1999`, `2000`, key = "year", value = "population")
left_join(tidy4a, tidy4b)
```

Now the data is tidy!

**Spread()**

Spreading is the opposite of gathering. For example, in `table2`: an observation is a country in a year, but each observation is spread across two rows.

```{r spreaddata}
table2
```

To tidy this up, we first analyse the represenation in similar way to `gather()`. This time, however, we only need two parameters:

* The column that contains variables names, the `key` column. Here, it's `type`.
* The column that contains values forms multiple variables, the `value` column. Here it's `count`.

```{r spread1}
spread(table2, key = type, value = count)
```

The functions `spread` and `gather` are not perfectly symmetrical because column type information is not transferred between them. The `convert` argument tries to convert character vectors to the appropriate type. Need to set `convert = TRUE`.

## 12.4 Separating and uniting

**Separate**

`separate()` pulls apart one column into multiple columns.

```{r separate1}
table3 %>%
  separate(rate, into = c("cases", "population"))
```

Can determine what makes the separationg with 

`sep = "/"`, or whatever the separator is.

By using `convert = TRUE`, `separate()` converts the data into better types.

**Unite**

`unite()` is the inverse of separate.

Can be used to rejoin data.

```{r unite1}
table5 %>%
  unite(new, century, year, sep = "")
```

## 12.5 Missing values

**Explicitly** flagged with `NA`
**Implicitly** not present in the data

```{r tidymissing}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)

# The way that a data set is represented can make implicit values explicit.

stocks %>% 
  spread(year, return)

# Because they might not be important. You can set na.rm = TRUE in gather().

stocks %>% 
  spread(year, return) %>% 
  gather(year, return, `2015`:`2016`, na.rm = TRUE)

# Another tool to make missing values explicit is complete()
stocks %>%
  complete(year, qtr)
```

## 12.6 Case Study

Dataset:

```{r casestudy1}
who
```

The best place to start is almost always to gather the columns that are not variables. 

* It looks like `country`, `iso2`, and `iso3` are three variables that redundantly specify the country.
* `year` is also clearly also a variable.
* We don't know what all the other columns are yet, but given the structure in the variable names (e.g. `new_sp_m014`, `new_ep_m014`, `new_ep_f014`) these are likely to be values, not variables.

Need to gather all the columns from `new_sp_m014` to `newrel_f65`. Give them the generic name `"key"` and use the variable `cases`. There are a lot of missing values, so for now we'll use `na.rm` so we can focus on the values that are present.

```{r casestudy2}
who1 <- who %>%
  gather(new_sp_m014:newrel_f65, key = "key", value = "cases", na.rm = TRUE)
who1
```

1. The first three letters of each column denote whether the column contains new or old cases of TB.
2. The next two letters describe the type of TB:
* `rel` stands for cases of relapse
* `ep` stands for cases of extrapulmonary TB
* `sn` stands for cases of pulmonary TB that could not be diagnosed by a pulmonary smear (negative)
* `sp` stands for cases of pulmonary TB that could not be diagnosed by a pulmonary smear (positive)
3. The sixth letter gives the sex of TB patiens.
4. The remaining numbers gives the age group.

Need to make a minor fix to the format of the column names as the names are inconsistent with `new_rel` we have `newrel`.

```{r casestudy3}
who2 <- who1 %>%
  mutate(key = stringr::str_replace(key, "newrel", "new_rel"))
who2
```

Now we can separate the values in each code with two passes of `separate()`.

```{r casestudy4}
who3 <- who2 %>%
  separate(key, c("new", "type", "sexage"), sep = "_")
who3

# Might as well drop the new column because it's constant.

who3 %>%
  count(new)

who4 <- who3 %>%
  select(-new, -iso2, -iso3)
```

Next, separate `sexage` into `sex` and `age`.

```{r casestudy5}
who5 <- who4 %>%
  separate(sexage, c("sex", "age"), sep = 1)
who5
```

Here's the code altogether:

```{r casestudy6}
who %>%
  gather(code, value, new_sp_m014:newrel_f65, na.rm = TRUE) %>% 
  mutate(code = stringr::str_replace(code, "newrel", "new_rel")) %>%
  separate(code, c("new", "var", "sexage")) %>% 
  select(-new, -iso2, -iso3) %>% 
  separate(sexage, c("sex", "age"), sep = 1)
```

How to confirm redundancy of `iso2` and `iso3`?

```{r casestudy7}
select(who3, country, iso2, iso3) %>%
  distinct() %>%
  group_by(country) %>%
  filter(n() > 1)
```