---
title: "Chapter 13"
author: "Miguel Arias"
date: "9/5/2017"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Relational data

```{r start, echo = FALSE}
library(tidyverse)
library(nycflights13)
```

Multiple tables of data are called **relational data** because it is the relations, not just the individual datasets, that are important.

Relational data verbs:

* **Mutating joins**, add new variables to one data from from matching observations in another
* **Filtering joins**, filter observations from one data frame based on whether or not they match an observation in the other table.
* **Set operations**, treat observations as if they were set elements.

## 13.2 nycflights13

The nycflights13 package contains four tibbles that are related to the `flights` table.

* `airlines` lets you loop up the full carrier name from its abbreviated code.
```{r nycfligths1}
airlines
```

* `airports` gives information about each airport, identified by the `faa` airport code:
```{r nycflights2}
airports
```

* `planes` gives information about each plane, identified by its `tailnum`
```{r nycflights3}
planes
```

* `weather` gives the weather at each NYC airport for each hour
```{r nycflights4}
weather
```

## 13.3 Keys

Two types of keys:
1. A **primary key** uniquely identifies an observation in its own table. For example, `planes$tailnum` is a primary key because it uniquely identifies each plane in the `planes` table.
2. A **foreign key** uniquely identifies an observation in another table. For example, the `flights$tailnum` is a foreign key because it appears in the `flights` table where it matches each flight to a unique plane.

Good practice to verify primary keys:
```{r keys1}
planes %>%
  count(tailnum) %>%
  filter(n > 1)

weather %>%
  count(year, month, day, hour, origin) %>%
  filter(n > 1)
```

### 13.3.1 Exercises

1. Add a surrogate key to flights

Add the column `flight_id` as a surrogate key.

```{r keysex}
flights %>%
  arrange(year, month, day, sched_dep_time, carrier, flight) %>%
  mutate(flight_id = row_number()) %>%
  glimpse()
```

## 13.4 Mutating Joins

Like `mutate()`, the join functions add variables to the right, so if you have a lot of variables already, the new variables won't get printed.

```{r mutatejoins1}
flights2 <- flights %>%
  select(year:day, hour, origin, dest, tailnum, carrier)
flights2

# If you want to add the full airline name to flights2
# Can combine the airlines and flights2 data frames with left_join()
flights2 %>%
  select(-origin, -dest) %>%
  left_join(airlines, by = "carrier")
```

**Types**

```{r jointypes}
x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     3, "x3"
)
y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2",
     4, "y3"
)
```

* `inner_join()`: matches pairs of observations whenever their keys are equal.
```{r innerjoin1}
x %>%
  inner_join(y, by = "key")
```

* `left_join()` keeps all observations of `x`
* `right_join()` keeps all observations of `y`
* `full_join()` keeps all observations in `x` and `y`.

### 13.4.5 Defining the key columns

* The default, `by = NULL`, uses all variables that appear in both tables, the so called **natural** join. For example, the flights and weather tables match on their common variables: `year`, `month`, `day`, `hour`, and `origin`.
```{r keycols1}
flights2 %>%
  left_join(weather)
```

* A character vector, `by = "x"`.
```{r keycols2}
flights2 %>%
  left_join(planes, by = "tailnum")
```

* A named character vector: `by = c("a" = "b")`. Matches variable `a` in table `x` to variable `b` in table `y`.

```{r keycols3}
flights2 %>%
  left_join(airports, c("dest" = "faa"))
```

### 13.4.6 Exercises

1. Compute the average delay by destination, then join on the airports data frame so you can show the spatial distribution of delays.

```{r keysex1}
library(purrr)
airports %>%
  semi_join(flights, c("faa" = "dest")) %>%
  ggplot(aes(lon, lat)) +
  borders("state") +
  geom_point() + 
  coord_quickmap()
```

```{r keysex1.1}
avg_dest_delays <-
  flights %>%
  group_by(dest) %>%
  # arrival delay NA's are cancelled flights
  summarise(delay = mean(arr_delay, na.rm = TRUE)) %>%
  inner_join(airports, by = c(dest = "faa"))

avg_dest_delays %>%
  ggplot(aes(lon, lat, colour = delay)) +
    borders("state") +
    geom_point() + 
    coord_quickmap()
```

2. Add the location of the origin and destination (i.e. the lat and lon) to flights

```{r keysex2}
flights %>%
  left_join(airports, by = c(dest = "faa")) %>%
  left_join(airports, by = c(origin = "faa")) %>%
  head()
```

3. Is there a relationship between the age of a plane and its delays?

```{r keysex3}
plane_ages <-
  planes %>%
  mutate(age = 2013 - year) %>%
  select(tailnum, age)

flights %>%
  inner_join(plane_ages, by = "tailnum") %>%
  group_by(age) %>%
  filter(!is.na(dep_delay)) %>%
  summarise(delay = mean(dep_delay)) %>%
  ggplot(aes(x = age, y = delay)) +
  geom_point()+
  geom_line()
```

## 13.5 Filtering joins

* `semi_join`: keeps all obs in `x` with a match in `y`. Useful for matching filtered summary tables back to the original rows.
* `anti_join`: drops all obs in `x` with a match in `y`. Useful for diagnosing mismatches.

For example, when connecting `flights` and `planes`, you might be interested to know that there are many `flights` that don't have a match in `planes`:
```{r filterjoins1}
flights %>%
  anti_join(planes, by = "tailnum") %>%
  count(tailnum, sort = TRUE)
```

### 13.5.1 Exercises

2. Filter flights to only show flights with planes that have flown at least 100 flights

```{r filterjoinex1}
planes_gt100 <- 
  filter(flights) %>%
  group_by(tailnum) %>%
  count() %>%
  filter(n > 100)

flights %>%
  semi_join(planes_gt100, by = "tailnum")
```

4. What does` anti_join(flights, airports, by = c("dest" = "faa"))` tell you? What does `anti_join(airports, flights, by = c("faa" = "dest"))` tell you?

`anti_join(flights, airports, by = c("dest" = "faa"))` are flights that go to an airport that is not in FAA list of destinations, likely foreign airports.

`anti_join(airports, flights, by = c("faa" = "dest"))` are US airports that don’t have a flight in the data, meaning that there were no flights to that aiport from New York in 2013.
  
## 13.7 Set operations

The final type of two-table verb are the set operations
* `intersect(x,y)`: return only observations in both `x` and `y`
* `union(x, y)`: return unique observations in `x` and `y`
* `setdiff(x, y)`: return observations in `x`, but not in `y`.