---
title: "R for Data Science"
author: "Miguel Arias"
date: "9/1/2017"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE, cache = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R For Data Science

## Data Visualisation

R has several systems for making graphs, but `ggplot2` is one of the most elegant and versatile. It implements the *grammar of graphics*, a coherent system for describing and building graphs.

In order to use `ggplot2`, we first need to load the `tidyverse` package.

```{r, results='hide'}
library(tidyverse)
```

1. Do cars with big engines use more fuel than cars with small engines? Use the `mpg` *data frame* found in ggplot2.

```{r mpg}
mpg
?mpg
```

Among the variables in `mpg` are:

1. `displ`, a car's engine size, in litres.
2. `hwy`, a car's fuel efficiency on the highway, in miles per gallon (mpg). A car with low fuel efficiency consumes more fuel than a car with high fuel efficiency when they travel the same distance.

### Creating a ggplot

To plot `mpg`:
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

### Graphing a template

Template for making graphs:

`ggplot(data = <DATA>) + <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))`
    
### Exercises

1. Run `ggplot(data = mpg)`. What do you see?

```{r}
ggplot(data = mpg)
```

There's an empty grey box.

2. How many rows are in `mpg`? How many columns?

```{r}
mpg
?mpg
```

There's 234 rows and 11 variables (columns).

3. What does `drv` variable describe?

`drv`: f = front-wheel drive, r = rear wheel drive, 4 = 4wd.

4. Make a scatterplot of `hwy` vs `cyl`

```{r scatterplot}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = hwy, y = cyl))
```

5. What happens if you make a scatter plot of `class` vs `drv`? Why is the plot not useful?

```{r trash}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = class, y = drv))
```
It makes a scatter plot, however there is not enough information for this to be useful. Only portrays if a vehicle is front-wheeler, rear-wheeler, or 4-wheeler.

## Aesthetic mappings


