---
title: "Chapter 4"
author: "Miguel Arias"
date: "9/5/2017"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Workflow 

```{r start, include = FALSE}
library(tidyverse)
library(ggplot2)
```

## 4.1 Coding basics

You can create new objects with `<-`:
```{r coding1}
x <- 3 * 4
```

All R statements where you create objects, **assignemnet** statements, have the same form:

`object_name <- value`

## 4.2 What's in a name?

Recommend using **snake_case** where you separete lowercase words with `_`.

This is how object/function names should look: `i_use_snake_case`

## 4.3 Calling functions

R has a large collection of built-in functions that are called like this:
`function_name(arg1 = val1, arg2 = val2, ...)`

To create and call out an object:
```{r coding2}
(y <- seq(1,10, length.out = 5))
```

## 4.4 Practice

1. Why does this code not work?

```{r codprac1}
my_variable <- 10
#my_varıable
```

It does not work because `my_variable` is misspelled the second time.

2. Tweak each of the following R commands so that they run correctly:

```{r codprac2}
library(tidyverse)

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))

filter(mpg, cyl == 8)
filter(diamonds, carat > 3)
```
