---
title: "Chapter 10"
author: "Miguel Arias"
date: "9/5/2017"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=6, fig.height=4)
```

# Tibbles

```{r start, include = FALSE}
library(tidyverse)
```

In order to coerce a data frame into a tibble. You can do that with `as_tibble()`:

```{r tibble1}
as_tibble(iris)
```

Can create a new tibble from individual vectors with `tibble()`.

```{r tibble2}
tibble(
  x = 1:5,
  y = 1,
  z = x ^ 2 + y
)
```

Another way to create a tibble is with `tribble()`.

```{r tribble1}
tribble(
  ~x, ~y, ~z,
  #--|--|----
  "a", 2, 3.6,
  "b", 1, 8.5
)
```

The comment line `#--|--|----` is useful to make it really clear where the header is.

## 10.3 Tibbles vs. data frame

**Printing**

Tibbles have a refined print method that shows only the first 10 rows, and all the columns that fit on screen. In addition to its name, each column reports its type, a nice feature borrowed from `str()`:

```{r tibble3}
tibble(
  a = lubridate::now() + runif(1e3) * 86400,
  b = lubridate::today() + runif(1e3) * 30,
  c = 1:1e3,
  d = runif(1e3),
  e = sample(letters, 1e3, replace = TRUE)
)
```

You can explictly `print()` the data frame and control the number of rows(`n`) and the `width` of the display. `width = Inf` displays all columns:

```{r printing1, results = 'hide'}
nycflights13::flights %>%
  print(n = 10, width = Inf)
```

### 10.3.2 Subsetting

If you want to pull out a single variable, you need some new tools, `$` and `[[`. `[[` can extract by name or position; `$` only extracts by name but is a little less typing.

```{r subsetting1}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

# Extract by name
df$x
df[["x"]]

# Extract by position
df[[1]]
```

Can be used in a pipe. Only need to use the special placeholder `.`:
```{r subestting2}
df %>% .$x
df %>% .[["x"]]
```

Some older functions don't work with tibbles. Need to use `as.data.frame()`

```{r asdataframe1}
#class(as.data.frame(tb))
```

## 10.5 Exercises

1. How can you tell if an object is a tibble?

A data frame will print the entire contents. A tibble will only print (by default) the first 10 rows and as many columns as will fit in the console.

2. Compare and contrast the following operations on a `data.frame` and equivalent tibble. What is different? Why might the default data frame behaviours cause you frustration?

```{r tibbleex2}
# on a data frame
df <- data.frame(abc = 1, xyz = "a")
df$x
df[, "xyz"]
df[, c("abc", "xyz")]

# on a tibble
df <- tibble(abc = 1, xyz = "a")
df$x
df[, "xyz"]
df[, c("abc", "xyz")]
```

* Tibbles never do partial matching; data frames do.
* Subsetting tibbles using `[[` will always return a tibble; subsetting data frames using `[[` can potentially return a vector.

3. If you have the name of a variable stored in an object, e.g. `var <- "mpg"`, how can you extract the reference variable from a tibble?

```{r tibbleex3}
var <- "hwy"
mpg[[var]]
```

4. Practice referring to non-systematic names in the following data frame by:

```{r tibbleex4}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)

# 1  Extracting the variable called `1`.
annoying %>% .$`1`

# 2 Plotting a scatterplot of `1` vs `2`.
ggplot(annoying, aes(`1`, `2`)) +
  geom_point()
# 3 Creating a new column called `3` which is `2` divided by `1`
(annoying <- mutate(annoying, `3` = `2` / `1`))
# 4 Renaming the columns to `one`, `two` and `three`
rename(annoying,
       one = `1`,
       two = `2`,
       three = `3`)
```

5. What does `tibble::enframe()` do? When might you use it?

`enframe()` is a helper function that converts named atomic vectors or lists to two-column data frames. You might use it if you have data stored in a named vector and you want to add it to a data frame and preserve both the name attribute and the actual value.

6. What option control how many additional column names are printed at the footer of a tibble?

`getOption("tibble.max_extra_cols")`
