---
title: "Chapter 15"
author: "Miguel Arias"
date: "9/5/2017"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=6, fig.height=4)
```

## Factors

```{r start, include = FALSE}
library(tidyverse)
library(forcats)
```

### 15.2 Creating factors

To create a factor you must start by creating a list of the valid **levels**:

```{r factors1}
x1 <- c("Dec", "Apr", "Jan", "Mar")
month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun",
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)
```

Now can create a factor:
```{r factors2}
y1 <- factor(x1, levels = month_levels)
y1
sort(y1)
```

The levels are needed, because if you omit them. They'll be taken from the data in alphabetical order:
```{r factors3}
factor(x1)
```

Sometimes prefer order of levels to match order of first appearance in data. Can do it by setting levels to `unique(x)`, or after the fact, with `fct_inorder()`:

```{r factors4}
f1 <- factor(x1, levels = unique(x1))
f1
f2 <- x1 %>% factor() %>% fct_inorder()
f2
```

If ever need to access the set of valid levels (`levels()`):
```{r factors5}
levels(f2)
```

### 15.3 General Social Survey

```{r gss1}
gss_cat
```

When factors are stored in a tibble, you can't see their levels so easily. One way to see them is with `count()`:

```{r gss2}
gss_cat %>%
  count(race)

# Or with a bar chart
ggplot(gss_cat, aes(race)) +
  geom_bar()

# By default, ggplot2 drops empty levels. Can force to display them with:
ggplot(gss_cat, aes(race)) +
  geom_bar() +
  scale_x_discrete(drop = FALSE)
```

#### 15.3.1 Exercise

1. Explore the distribution of `rincome` (reported income). What makes the default bar chart hard to understand? How could you improve the plot?

```{r factorex1}
rincome_plot <-
  gss_cat %>%
  ggplot(aes(rincome)) +
  geom_bar()
rincome_plot
```

The labels are too close to read. Could change the angle of the labels or use `coord_flip`.

```{r factorex1.1}
rincome_plot +
  theme(axis.text.x = element_text(angle = 90))
rincome_plot +
  coord_flip()
```

2. What is the most common `relig` in this survey? What's the most common `partyid`?

```{r factorex2}
gss_cat %>%
  count(relig) %>%
  arrange(-n) %>%
  head(1)

gss_cat %>%
  count(partyid) %>%
  arrange(-n) %>%
  head(1)
```

3. Which `relig` does `denom` (denomination) apply to? How can you find out with a table? How can you find out with a visualisation?

```{r factorex3}
levels(gss_cat$denom)

# Denom refers to "Protestant"
gss_cat %>%
  filter(!denom %in% c("No answer", "Other", "Don't know", "Not applicable",
                       "No denomination")) %>%
  count(relig)

# Can see it clearly in a scatterplot
gss_cat %>%
  count(relig, denom) %>%
  ggplot(aes(x = relig, y = denom, size = n)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90))
```

### 15.4 Modifying factor order

Often useful to change the order of the factor levels in a visualisation. For example, imagine you want to explore the average number of hours spent watching TV per day across religions:

```{r factororder1}
relig_summary <- gss_cat %>%
  group_by(relig) %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )

ggplot(relig_summary, aes(tvhours, relig)) + geom_point()

# Here can use fct_reorder
# f, the factor whose levels you want to modify
# x, a numeric vector that you want to use to reoder the levels
ggplot(relig_summary, aes(tvhours, fct_reorder(relig, tvhours))) +
  geom_point()

# By reordering religion it is easier to see who watches more TV
```

As the transformations become more complicated. Better to move them out of `aes()` and into a separate `mutate()` step. Can rewrite the plot above as:

```{r factororder2}
relig_summary %>%
  mutate(relig = fct_reorder(relig, tvhours)) %>%
  ggplot(aes(tvhours, relig)) +
    geom_point()
```

**Only use `fct_reorder()` for factors that are arbitrarily ordered**

Can use `fct_relevel()`, as it takes a factor (`f`) and then any number of levels that you want to move to the front of the line.

```{r factororder3}
rincome_summary <- gss_cat %>%
  group_by(rincome) %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )
ggplot(rincome_summary, aes(age, fct_reorder(rincome, age))) + geom_point()

# Here fct_reorder does not make much sense
# Better to use fct_relevel
ggplot(rincome_summary, aes(age, fct_relevel(rincome, "Not applicable"))) +
  geom_point()
```

`fct_reorder2()` reorders the factor by the `y` values associated with the largest `x` values. This makes the plot easier to read because the line colours line up with the legend.

```{r factororder4}
by_age <- gss_cat %>%
  filter(!is.na(age)) %>%
  group_by(age, marital) %>%
  count() %>%
  mutate(prop = n / sum(n))

ggplot(by_age, aes(age, prop, colour = marital)) +
  geom_line(na.rm = TRUE)

ggplot(by_age, aes(age, prop, colour = fct_reorder2(marital, age, prop))) +
  geom_line() +
  labs(colour = "marital")
```

For bar plots, can use `fct_infreq()` to order levels in increasing frequency (doesn't need variables). May want to combine it with `fct_rev()`

```{r factororder5}
gss_cat %>%
  mutate(marital = marital %>% fct_infreq() %>% fct_rev()) %>%
  ggplot(aes(marital)) +
    geom_bar()
```

#### 15.4.1 Exercises

1. There are some suspiciously high numbers in `tvhours`. Is the `mean` a good summary?

```{r fctorderex1}
summary(gss_cat[["tvhours"]])
gss_cat %>%
  filter(!is.na(tvhours)) %>%
  ggplot(aes(x = tvhours)) +
  geom_histogram(binwidth = 1)
```

2. For each factor in `gss_cat` identify whether the order of the levels is arbitrary or principled.

```{r fctorderex2}
# Prints the names of all the factors
keep(gss_cat, is.factor) %>% names()
```

Six categorical values: `marital`, `race`, `rincome`, `partyid`, `relig`, `denom`.

3. Why did moving "Not applicable" to the front of the levels move it to the bottom of the plot?

Because it gives the level "Not applicable" an integer value of 1.

### 15.5 Moifying factor levels

`fct_recode()` will leave levels that aren't explicitly mentioned as is, and will warn you if you accidentally refer to a level that doesn't exist.

```{r fctmodify1}
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
    "Republican, strong"    = "Strong republican",
    "Republican, weak"      = "Not str republican",
    "Independent, near rep" = "Ind,near rep",
    "Independent, near dem" = "Ind,near dem",
    "Democrat, weak"        = "Not str democrat",
    "Democrat, strong"      = "Strong democrat",
    "Other"                 = "No answer",
    "Other"                 = "Don't know",
    "Other"                 = "Other party"
  )) %>%
  count(partyid)
```

* `fct_collapse()` is a useful variant of `fct_recode()`. For each new variable, can provide a vector of old levles.
* `fct_lump()` allows you to lump together all the small groups to make a plot or table simpler.
- Can use `n` parameter to specify how many groups (excluding other) you want to keep.

#### 15.5.1 Exercises

1. How have the proportions of people identifying as Democrat, Republican, and Independent changed over time?

Need to combine the multiple levels into Democrat, Republican, and Independent
```{r fctmodex1}
levels(gss_cat$partyid)
gss_cat %>% 
  mutate(partyid = 
           fct_collapse(partyid,
                        other = c("No answer", "Don't know", "Other party"),
                        rep = c("Strong republican", "Not str republican"),
                        ind = c("Ind,near rep", "Independent", "Ind,near dem"),
                        dem = c("Not str democrat", "Strong democrat"))) %>%
  count(year, partyid)  %>%
  group_by(year) %>%
  mutate(p = n / sum(n)) %>%
  ggplot(aes(x = year, y = p,
             colour = fct_reorder2(partyid, year, p))) +
  geom_point() +
  geom_line() +
  labs(colour = "Party ID.")
```

2. How could you collapse `rincome` into a small set of categories?

Group all the non-response into one category, and then group other categories into a smaller number.

```{r fctmodex2}
levels(gss_cat$rincome)
library("stringr")
gss_cat %>%
  mutate(rincome =
           fct_collapse(
             rincome,
             `Unknown` = c("No answer", "Don't know", "Refused", "Not applicable"),
             `Lt $5000` = c("Lt $1000", str_c("$", c("1000", "3000", "4000"),
                                              " to ", c("2999", "3999", "4999"))),
             `$5000 to 10000` = str_c("$", c("5000", "6000", "7000", "8000"),
                                      " to ", c("5999", "6999", "7999", "9999"))
           )) %>%
  ggplot(aes(x = rincome)) +
  geom_bar() + 
  coord_flip()
```

