---
title: "Chapter 21"
author: "Miguel Arias"
date: "9/6/2017"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=6, fig.height=4)
```

## Iteration

```{r start, include = FALSE}
library(tidyverse)
```

### 21.2 For loops

We have this simple tibble:
```{r loops1}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

# We want to compute the median of each col:
median(df$a)
median(df$b)
median(df$c)
median(df$d)
```

Can use a loop:

```{r loops2}
output <- vector("double", ncol(df))  # 1. output
for (i in seq_along(df)) {            # 2. sequence
  output[[i]] <- median(df[[i]])      # 3. body
}
output
```

Every loop has three components:
1. The **output**: `output <- vector("double", length(x))`. Need to allocate sufficient space for the output.

2. The **sequence**: `i in seq_along(df)`. This determines what to loop over: each run of the for loop will assign `i` to a different value from `seq_along(df)`. It's useful to think of `i` as a pronoun, like "it".

`seq_along()` is a safe version of `1:length(l)`.

3. The **body**: `output[[i]] <- median(df[[i]])`. The code that does the work. It's run repeatedly, each time with a different value for `i`. 

#### 21.2.1 Exercises

1. Write for loops to:

```{r loopsex1}
# Compute the mean of every column in mtcars
output <- vector("double", ncol(mtcars))
names(output) <- names(mtcars)
for (i in names(mtcars)) {
  output[i] <- mean(mtcars[[i]])
}
output

# Determine the type of each column in nycflights13::flights
data("flights", package = "nycflights13")
output <- vector("list", ncol(flights))
names(output) <- names(flights)
for (i in names(flights)) {
  output[[i]] <- class(flights[[i]])
}
output

# Compute the number of unique values in each column of iris
data(iris)
iris_uniq <- vector("double", ncol(iris))
names(iris_uniq) <- names(iris)
for (i in names(iris)) {
  iris_uniq[i] <- length(unique(iris[[i]]))
}
iris_uniq

# Generate 10 random normals for each mu = -10, 0, 10, and 100

# number to draw
n <- 10
# values of the mean
mu <- c(-10, 0, 10, 100)
normals <- vector("list", length(mu))
for (i in seq_along(normals)) {
  normals[[i]] <- rnorm(n, mean = mu[i])
}
normals
```

### 21.3 For loop variations

1. Modifying an existing object, instead of creating a new object.
2. Looping over names or values, instead of indices.
3. Handling outputs of unknown length.
4. Handling sequences of unknown length.

Better to use `[[` in for loops instead of `[`. As it makes it clear that I want to work with a single element.

#### 21.3.2 Looping partners

Three basic ways to loop over a vector.

1. Most general: `for (i in seq_along(xs))`, and extracting the value with `x[[i]]`.

2. Looping over the elements: `for (x in xs)`. Most useful if only care about plotting or saving a file. 

3. Looping over the names: `for (nm in names(xs))`. This gives you the name, which you can use to access the value with `x[[nm]]`. Useful if want to use the name in a plot title or a file name. 

If creating an output, make sure to name the results vector like:
`results <- vector("list", length(x))`
`names(results) <- names(x)`

#### 21.3.3 Unknown output length

Save results in a list, and then combine into a single vector.

```{r loops3}
means <- c(0, 1, 2)

out <- vector("list", length(means))
for (i in seq_along(means)) {
  n <- sample(100, 1)
  out[[i]] <- rnorm(n, means[[i]])
}
str(out)
str(unlist(out))
```

This occurs in other places:
* Might be generating a long string. Instead of `paste()`ing together each iteration with the previous, save the output in a character vector and then combine that vector in a single string with `paste(output, collaspe = "")`.
* Might be generating a big data frame. Instead of sequentially `rbind()`ing in each iteration, save the output in a list, then use `dplyr::bind_rows(output)` to combine the output into a single data frame.

#### 21.3.4 Unknown seq length

Use a while loop

```{r while}
# while (condition) {
    #  body
# }

# How many tries it takes to get three heads in a row:
flip <- function() sample(c("T", "H"), 1)

flips <- 0
nheads <- 0

while (nheads < 3) {
  if (flip() == "H") {
    nheads <- nheads + 1
  } else {
    nheads <- 0
  }
  flips <- flips + 1
}
flips
```

#### 21.3.5 Exercises

1. Imagine you have a directory full of CSV files that you want to read in. You have their paths in a vector, `files <- dir("data/", pattern = "\\.csv$", full.names = TRUE)`, and now you want to read each one with `read_csv()`. Write the for loop that will load them into a single data frame.

```{r loopsex2}
files <- dir("data/", pattern = "\\.csv$", full.names = TRUE)

# pre-allocate a list
df <- vector("list", length(files))
# Create list of data frames
for (fname in seq_along(files)) {
  df[[i]] <- read_csv(files[[i]])
}
# Use bind_rows to create a single data frame
df <- bind_rows(df)
```

2. What happens if you use `for (nm in names(x))` and `x` has no names? What if only some of the elements are named?

If there are no names for the vector, it does not run the code in the loop. When it reaches an unnamed element it produces an error.

3. Write a function that prints the mean of each numeric column in a data frame, along with its name.

```{r loopsex3}
library(stringr)
show_mean <- function(df, digits = 2) {
  # Get max length of any variable in the dataset
  maxstr <- max(str_length(names(df)))
  for (nm in names(df)) {
    if (is.numeric(df[[nm]])) {
      cat(str_c(str_pad(str_c(nm, ":"), maxstr + 1L, side = "right"),
                format(mean(df[[nm]]), digits = digits, nsmall = digits),
                sep = " "),
          "\n")
    }
  }
}
show_mean(iris)
```

4. What does this code do? How does it work?

```{r loopsex4}
trans <- list( 
  disp = function(x) x * 0.0163871,
  am = function(x) {
    factor(x, labels = c("auto", "manual"))
  }
)

for (var in names(trans)) {
  mtcars[[var]] <- trans[[var]](mtcars[[var]])
}
```

This code mutates the `disp` and `am` columns:
* `disp` is multiplied by 0.0163871
* `am` is replaced by a factor variable

The code works by looping over a named list of functions. It calls the named function in the list on the column of `mtcars` with the same name, and replaces the values of that column.

### 21.4 For loops vs. functionals

Can create functions to run for loops and prevent repetition and a chance for bugs. Also, can generalise these functions in order to compute different arguments, like `col_mean()`, `col_median()` and `col_sd()`. 

For example:
```{r loops4}
col_summary <- function(df, fun) {
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[i] <- fun(df[[i]])
  }
  out
}
col_summary(df, median)
col_summary(df, mean)
```

#### 21.4.1 Exercises

1. Read the documentation for `apply()`. In the 2d case, what two for loops does it generalise.

Looping over the rows or columns of a matrix or data frame.

2. Adapt `col_summary()` so that it only applies to numeric columns. Might want to use `is_numeric()` function that returns a logical vector that has a `TRUE` corresponding to each column.

```{r loopsex5}
col_summary2 <- function(df, fun) {
  # test whether each colum is numeric
  numeric_cols <- vector("logical", length(df))
  for (i in seq_along(df)) {
    numeric_cols[[i]] <- is.numeric(df[[i]])
  }
  # indexes of numeric columns
  idxs <- seq_along(df)[numeric_cols]
  # number of numeric columns
  n <- sum(numeric_cols)
  out <- vector("double", n)
  for (i in idxs) {
    out[i] <- fun(df[[i]])
  }
  out
}
```

### 21.5 The map functions

* `map()` makes a list
* `map_lgl()` makes a logical vector
* `map_int()` makes an integer vector
* `map_dbl()` makes a double vector
* `map_chr()` makes a character vector

Can use these functions to perform the same computations as the last for loop.
```{r loops5}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

map_dbl(df, mean)
map_dbl(df, median)
map_dbl(df, sd)
```

#### 21.5.2 Base R

* `lapply()` is basically identical to `map()`, except that `map()` is consistent with all the other function in purrr, and you can use the shortcuts for `.f`.
* Base `sapply()` is a wrapper around `lapply()` that automatically simplifies the output. This is useful for interactive work but is problematic in a function because you never know what sort of output you'll get:

```{r base1}
x1 <- list(
  c(0.27, 0.37, 0.57, 0.91, 0.20),
  c(0.90, 0.94, 0.66, 0.63, 0.06), 
  c(0.21, 0.18, 0.69, 0.38, 0.77)
)
x2 <- list(
  c(0.50, 0.72, 0.99, 0.38, 0.78), 
  c(0.93, 0.21, 0.65, 0.13, 0.27), 
  c(0.39, 0.01, 0.38, 0.87, 0.34)
)

threshold <- function(x, cutoff = 0.8) x[x > cutoff]
x1 %>% sapply(threshold) %>% str()
x2 %>% sapply(threshold) %>% str()
```

* `vapply()` is a safe alternative to `sapply()`, but it's a lot of typing.

#### 21.5.3 Exercises

1. Write code that uses one of the map functions to:

```{r mapex1}
# Compute the mean of every column in `mtcars`
map_dbl(mtcars, mean)

# The type of every column in `nycflights13::flights`
map(nycflights13::flights, class)
map_chr(nycflights13::flights, typeof)

# The number of unique values in each column of `iris`
map_int(iris, ~ length(unique(.)))

# Generates 10 random normals for each of mu = -10, 0, 10, and 100
map(c(-10, 0, 10, 100), rnorm, n = 10)
```

2. How can you create a single vector that for each column in a data frame indicates whether or not it's a factor?

Use `map_lgl` with the function `is.factor`.
```{r mapex2}
map_lgl(mtcars, is.factor)
```

3. What happens when you use the map functions on vectors that aren't lists? What does `map(1:5, runif)` do? Why?
The function `map` applies the function to each element of the vector.

```{r mapex3}
map(1:5, runif)
```

4. What does `map(-2:2, rnorm, n = 5)` do? Why? What does `map_dbl(-2:2, rnorm, n = 5)` do? Why?

This takes samples of `n = 5` for normal distributions of means -2, -1, 0, 1, and 2, an returns a list with each element a numeric vector of length 5.

```{r mapex4}
map(-2:2, rnorm, n = 5)

# If we use map_dbl, it produces an error

# Can use flatten_dbl
flatten_dbl(map(-2:2, rnorm, n = 5))
```

5. Rewrite `map(x, function(df) lm(mpg ~ wt, data = df))` to eliminate the anonymous function.

```{r mapex5}
map(list(mtcars), ~ lm(mpg ~ wt, data = .))
```

### 21.6 Dealing with failure

Can use:
* `safely()`: takes a function (a verb) and returns a modified version.
* `possibly()`: always succeeds, just like `safely()`. Give it a default value to return when an error occurs.
* `quietly()`: same as `safely()` but instead of capturing errors, it captures printed output, messages, and warnings.

### 21.7 Mapping over multiple arguments

`map2()` iterates over two vectors in parallel:

```{r map2}
mu <- list(5, 10, -3)
sigma <- list(1, 5, 10)
map2(mu, sigma, rnorm, n = 5) %>% str()
```

Just like `map()` it is just a wrapper around a for loop:

```{r map2.1}
map2 <- function(x, y, f, ...) {
  out <- vector("list", length(x))
  for (i in seq_along(x)) {
    out[[i]] <- f(x[[i]], y[[i]], ...)
  }
  out
}
```

There's also `pmap()` which takes a list of arguments.
```{r pmap}
n <- list(1, 3, 5)
args1 <- list(n, mu, sigma)
args1 %>%
  pmap(rnorm) %>% 
  str()

# Since the args are the same lenght it makes sense 
# to store them in a data frame:

params <- tribble(
  ~mean, ~sd, ~n,
    5,     1,  1,
   10,     5,  3,
   -3,    10,  5
)

params %>% 
  pmap(rnorm)
```


#### 21.7.1 Invoking different functions

By using `invoke_map()` we can apply different functions to different arguments simultaneously.

For example:
```{r invoke_map}
f <- c("runif", "rnorm", "rpois")
param <- list(
  list(min = -1, max = 1), 
  list(sd = 5), 
  list(lambda = 10)
)

invoke_map(f, param, n = 5) %>% str()
```

The first argument is a list of functions or character vector of function names. The second argument is a list of lists giving the arguments that vary for each function. The subsequent arguments are passed on to every function.

Can use `tribble()` to make creating these matching pairs a little easier:

```{r invoke_map2}
sim <- tribble(
  ~f,      ~params,
  "runif", list(min = -1, max = 1),
  "rnorm", list(sd = 5),
  "rpois", list(lambda = 10)
)
sim %>% 
  mutate(sim = invoke_map(f, params, n = 10))
```

#### 21.9.1 Predictable functions

* `keep()`: keeps elements of the input where the predicate is `TRUE`.
* `discard()`: keeps elements of the input where the predicate is `FALSE`.
* `some()`: determines if the predicate is true for any of the elements
* `every()`: determines if the predicate is true for all of the elements.
* `detect()`: finds the first element where the predicate is true; `detect_index()` returns its position.
* `head_while()` and `tail_while()` takes elements from the start or end of a vector while a predicate is true.

#### 21.9.2 Reduce and accumulate

* `reduce()`: used to reduce lists/vectors/data frames into a single one. For example:

```{r reduce}
# List
dfs <- list(
  age = tibble(name = "John", age = 30),
  sex = tibble(name = c("John", "Mary"), sex = c("M", "F")),
  trt = tibble(name = "Mary", treatment = "A")
)

dfs %>% reduce(full_join)

# Vectors
vs <- list(
  c(1, 3, 5, 6, 10),
  c(1, 2, 3, 7, 8, 10),
  c(1, 2, 3, 4, 8, 9, 10)
)

vs %>% reduce(intersect)
```

* `accumulate()`: similar to reduce but keeps all the interim results.

#### 21.9.3 Exercises

1. Implement your own version of `every()` using a for loop. Compare it with `purrr::every()`. 

```{r ex1}
# Use ... to pass arguments to the function
every2 <- function(.x, .p, ...) {
  for (i in .x) {
    if (!.p(i, ...)) {
      # If any is FALSE we know not all of then were TRUE
      return(FALSE)
    }
  }
  # if nothing was FALSE, then it is TRUE
  TRUE  
}

every2(2:4, function(x) {x > 3})
every2(2:4, function(x) {x > 1})
```

2. Create an enhanced `col_sum()` that applies a summary function to every numeric column in a data frame.

```{r ex2}
col_sum2 <- function(df, f, ...) {
  map(keep(df, is.numeric), f, ...)
}

col_sum2(iris, mean)
```